---
title: "ENU R workshop example script"
author: "Michelle Jamieson & Jan Savinc"
date: "`r Sys.Date()`"
output: 
  html_document:
    number_sections: true
    toc: true
editor_options:
	chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction to example script

This script contains examples of code presented in the ENU R workshop. The examples are intended to illustrate the concepts and methods in the presentation. They are designed to be run in succession - if you want to jump ahead to a particular section make sure to run the preceding scripts (in RStudio this can be done by clicking on the `Run All Chunks Above` button in the section you want to run from).


# Installing & loading packages

We start by installing the package `{pacman}` (package manager) the 'traditional' way. We then use the `{pacman}` package to install the other packages required.

```{r}
install.packages("pacman")  # the name of the package needs to be provided in quotes

# TODO: Michelle, will you put in code on how to install packages using pacman?
```

# Data frames / tibbles

We'll use the `starwars` tibble included in the {tidyverse} package to show how data frames / tibbles work.

```{r}
require(tidyverse)  # making sure that tidyverse is loaded

starwars  # calling the name of a data frame will print it...
print(starwars)  # ...the same as using print(df)
```

## Indexing / addressing values in data frames

```{r}
starwars[1, 4]  # this refers to the 1st row of the 4th column (which character does it refer to?)

starwars[3, ]  # this refers to 3rd row, all columns
starwars[, 6]  # this refers to all rows, column 6

starwars[,]  # this is the same as simply calling starwars. Give it a try!


starwars$height  # returns the 'height' column
starwars[,2]  # returns the same, but as a tibble with a single column - subtle difference!
starwars[,"height"]  # this also returns the 'height' column


## You can do more complicated things using vector notation as well, e.g.:

starwars[4:6, ]  # returns rows 4 to 6 and all columns

starwars[4:6, c(1,2,5,6,8,9)]  # returns rows 4 to 6 and the columns specified in c(...)

```

