---
title: "ENU R workshop example script"
author: "Michelle Jamieson & Jan Savinc"
date: "`r Sys.Date()`"
output: 
  html_document:
    number_sections: true
    toc: true
editor_options:
	chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction to example script

This script contains examples of code presented in the ENU R workshop. The examples are intended to illustrate the concepts and methods in the presentation. They are designed to be run in succession - if you want to jump ahead to a particular section make sure to run the preceding scripts (in RStudio this can be done by clicking on the `Run All Chunks Above` button in the section you want to run from).


# Installing & loading packages

We start by installing the package `{pacman}` (package manager) the 'traditional' way. We then use the `{pacman}` package to install the other packages required.

```{r}
install.packages("pacman")  # the name of the package needs to be provided in quotes

# TODO: Michelle, will you put in code on how to install packages using pacman?
```

