---
title: "ENU R workshop example script"
author: "Michelle Jamieson & Jan Savinc"
date: "`r Sys.Date()`"
output: 
  html_document:
    number_sections: true
    toc: true
editor_options:
	chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction to example script

This script contains examples of code presented in the ENU R workshop. The examples are intended to illustrate the concepts and methods in the presentation. They are designed to be run in succession - if you want to jump ahead to a particular section make sure to run the preceding scripts (in RStudio this can be done by clicking on the `Run All Chunks Above` button in the section you want to run from).


# Installing & loading packages

We start by installing the package `{pacman}` (package manager) the 'traditional' way. We then use the `{pacman}` package to install the other packages required.

```{r}
install.packages("pacman")  # the name of the package needs to be provided in quotes

library("pacman") # Load pacman package
# After running the previous lines of code, we can use the functions provided by the pacman package.

p_load(ggplot2, dplyr, stringr) # Install & load packages
# Note that the p_load function also checks whether a package is installed already. If the package is not installed yet, it is installed automatically by the pacman package.

# The p_load function basically replaces the Base R library, install.packages, and require functions and allows installing and loading numerous packages simultaneously using much less R code.

p_unload(ggplot2, dplyr, stringr)  # Unload packages
# The following packages have been unloaded:ggplot2, dplyr, stringr

p_update(update = FALSE) # Check for outdated packages

```

# Data frames / tibbles

We'll use the `starwars` tibble included in the {tidyverse} package to show how data frames / tibbles work.

```{r}
require(tidyverse)  # making sure that tidyverse is loaded

starwars  # calling the name of a data frame will print it...
print(starwars)  # ...the same as using print(df)
```

## Indexing / addressing values in data frames

```{r}
starwars[1, 4]  # this refers to the 1st row of the 4th column (which character does it refer to?)

starwars[3, ]  # this refers to 3rd row, all columns
starwars[, 6]  # this refers to all rows, column 6

starwars[,]  # this is the same as simply calling starwars. Give it a try!


starwars$height  # returns the 'height' column
starwars[,2]  # returns the same, but as a tibble with a single column - subtle difference!
starwars[,"height"]  # this also returns the 'height' column


## You can do more complicated things using vector notation as well, e.g.:

starwars[4:6, ]  # returns rows 4 to 6 and all columns

starwars[4:6, c(1,2,5,6,8,9)]  # returns rows 4 to 6 and the columns specified in c(...)

```


## Reading & saving data frames

```{r}
require(tidyverse)

## you can read from a file on your computer, or from the internet - in this
## example we will read a file from the National Records Scotland website
most_common_surnames_2022 <- read_csv(file = "https://www.nrscotland.gov.uk/files//statistics/common-surnames/most-com-surnames-bmd-registers-22.csv")
most_common_surnames_2022  # print the first few lines of the data

## this saves the starwars data frame as a CSV file (comma-separated values)
## inside your current project directory
starwars %>%
  write_csv(file = "./starwars.csv")
```



# Reading excel files

```{r}
require(readxl)  # the readxl package is needed

## in the below section we download an excel file to a temporary location

sample_excel_file <- tempfile()

download.file(
  url = "https://www.nrscotland.gov.uk/files/statistics/population-estimates/mid-21/mid-year-pop-est-21-data.xlsx",
  destfile = sample_excel_file,
  mode = "wb")

excel_sheets(path = sample_excel_file)  # show the names of the sheets/tabs in an excel file


## the two lines below read from the same sheet - you can use the name or its number
read_excel(path = sample_excel_file, sheet = 5)
read_excel(path = sample_excel_file, sheet = "Table 2")

## Ntoe how there's a lot of 'cruft' in the excel file because a lot of the
## cells are devoted to formatting and describing the data

## One approach to clean up the data would be to remove columns and rows that
## don't contain substantial information; however, we can also specify a range
## of cells to read from!
read_excel(
  path = sample_excel_file,
  sheet = "Table 2",
  range = "A5:G20",
  col_names = TRUE
)  # this reads from the 'rectangle' between A5 and G20, and also knows to treat the first row as column names
```


# Data cleaning

## {janitor} package examples

For this we will use the excel file from above.

```{r}
# TODO: find a good example of a dataset that can be used with janitor nicely 
# read_excel(path = sample_excel_file, sheet = 5)
```

## Renaming columns

```{r}
## using the tidyverse 'rename' function, we can specify which columns to rename
starwars %>%
  rename(height_in_cms = height, mass_in_kg = mass, skin_colour = skin_color)

## you can use 'select' if you want to rename and select only a subset of the
## columns
starwars %>%
  select(character_name = name, character_eye_colour = eye_color)

## you can name columns using the "names" function and assigning values to it,
## as well - this might look a bit strange because we are modifying the result
## of a function
names(starwars)  # this will print the column names

small_starwars <- starwars %>% select(1:5)
names(small_starwars) <- c("character_name", "character_height", "character_mass")  # we get a warning that the we did not name all of the columns
small_starwars  # two of the columns now have missing names
names(small_starwars) <- c("character_name", "character_height", "character_mass", "character_hair_colour", "character_eye_colour")
small_starwars  # now all the columns have names

## programmatically renaming columns - to do this, we need to use
## 'rename_with()'. For example, we can remove the keyword "character" from the
## names above
small_starwars %>%
  rename_with(~str_remove_all(string = .x, pattern = "character"))
## what we actually want is to remove the underscore as well
small_starwars %>%
  rename_with(~str_remove_all(string = .x, pattern = "character_"))

```


